CUDA-accelerated Genetic Feedforward-ANN Training for Data Mining

### 2010-04-21 doc/report.tex
SIGNING OFF HOPEFULLY NOT FOR THE WHOLE NIGHT, GOOD LUCK
git-svn-id: http://svn.scesoc.ca/evolution/trunk@114 fee98b4c-d1c3-4d34-9525-a468159e4a48

Compared to a traditional sequential implementation of GA training of ANNs, the use of graphical processing units (GPUs) as parallel processors provides significant performance improvements at a fraction of the price of alternatives such as cluster computing. GPUs, however, use a specialized programming paradigm which must be taken into account to leverage their full processing power. In this paper, we show an implementation of GA training of ANNs which achieves an order-of-magnitude speedup over a sequential algorithm.

@@

% ----------------------------------------------------------------------------
\subsection{Artificial Neural Networks} \label{ann}
% ----------------------------------------------------------------------------
ANNs are composed of nodes connected by weighted directed edges. In a feed-forward network architecture, nodes are organized into an input layer, one or more hidden layers, and an output layer. Each layer is fully connected to the next. Each of the hidden nodes at hidden layer $n$ perform a transcendental function with inputs equal to the outputs of layer $n-1$ each multiplied by the weight associated with the edge through which they are fed from layer $n-1$ to layer $n$. In a radial basis function (RBF) network, the nodes in each hidden layer calculate the distance between their inputs and a centre vector, and pass a weighted distance to a RBF. Nodes in a sigmoid function network perform a similar operation with a sigmoid function in place of the RBF. The output node is typically linear, computing a weighted sum of its inputs. The output of the ANN is a single real value which is interpreted as a classification confidence. ANN computation is dominated by the number of multiplications at the directed edges, proportional to the number of features in the data set and the number of hidden nodes, and by the transcendental functions at the hidden nodes, which are typically exponentiations.  ANN training has been implemented previously on GPU hardware, however these efforts have focused on training through backpropagation \cite{backprop} as opposed to genetic algorithms.

@@

% ############################################################################
\subsection{Classifier Design} \label{design}
% ############################################################################

While our GA-ANN classifier was designed for the 2009 UC San Diego Data Mining Contest data set, it is capable of adapting to a variety of data sets.  We have developped a flexible ANN topology, allowing for one or two hidden layers and an arbitrary number of nodes per hidden layer (within the limits of GPU memory). In addition, each layer of the ANN is capable of performing RBF or sigmoid calculations.

The classifier receives as inputs real-valued features which are standardized prior to processing by the ANN.  The classifier outputs a single real value, in the range of 0 and 1, which represents the confidence with which the pattern is a member of the positive class.  These confidence values can be used for ranking during lift calculations, or for sensitivity or specificity measures with an arbitrary value chosen as a threshold for positive identification of a pattern.  Therefore, the fitness function is not limited to lift at 20\% as presented in this paper.

In our final design, each candidate ANN is represented by a chromosome which consists of four arrays of real-valued numbers. Edge weights between input nodes and hidden layers, edge weights between hidden and output layers, width values for RBFs at hidden nodes and center values for RBFs at hidden nodes are all encoded, and can take on any real value. Figure~\ref{fig:evolution-lift} demonstrates the chromosome structure and its relation to ANN structure.  This structure is simple yet powerful; individual edges are 'removed' by the genetic algorithm when their weight is set to zero and nodes are 'removed' by the genetic algorithm when their RBF width value is set to zero.  Thus, we achieve the performance gains of a fixed ANN topology (allowing for disciplined memory access) while searching a range of topologies.

@@

% ----------------------------------------------------------------------------
\subsection{Data Preprocessing Steps} \label{preprocessing}
% ----------------------------------------------------------------------------
In order to handle nominal feature values in data sets, we have implemented an algorithm which, given a tab seperated values file, converts nominal feature values in the data set into numeric values.  The algorithm replaces each nominal value in the data set with the logarithm of the probability ratio between the positive and negative values observed given the nominal value.  As a result, meaningful numeric values can be extracted from nominal values without domain knowledge relating to the values. This improves on the traditional orthogonal coding method \cite[slide 17]{MLP-NominalCoding} by avoiding allocation of a disproportionate number of input nodes to nominal features.

@@

When applied to the 2009 UC San Diego Data Mining Contest blind test data set, the resulting classifier produced a lift at 20\% value of 4.031, placing 35th in the world for classification of this data set. Due to the stochastic nature of genetic algorithms and the contest rules which state that contestants may test any number of candidate classifiers, it is likely that repeated entry into this contest would result in a higher ranking. The discrepency between the estimated performance and actual performance on the blind test data set can be attributed to the following:
\begin{enumerate}
	\item Discrepency in test set size: the estimated performance measurements were done on sets of approximately 14000 patterns while the blind test data set has approximately 36000 patterns.
	\item Discrepency in class balance: when splitting the data set into training and hold-out data sets, class balance was not enforced for either of the sets. The blind test data set was known to have the same class imbalance as the given data set. Therefore, it is possible that the data used to collect experimental data had a class imbalance which did not reflect the blind test data set.
	\item Nominal values unique to the blind test data set: It is possible that patterns with nominal values that do not appear in the training data set could cause misclassification, however, the pre-processing step which converts nominal values into numeric values accounts for values which have not been seen previously.
\end{enumerate}

>>>>

Compared to a traditional sequential implementation of GA training of ANNs, the use of graphical processing units (GPUs) as parallel processors provides significant performance improvements at a fraction of the price of alternatives such as cluster computing. GPUs, however, use a specialized programming paradigm which must be taken into account to leverage their full processing power. In this report, we show an implementation of GA training of ANNs which achieves an order-of-magnitude speedup over a sequential algorithm.

@@

% ----------------------------------------------------------------------------
\subsection{Artificial Neural Networks} \label{ann}
% ----------------------------------------------------------------------------
ANNs are composed of nodes connected by weighted directed edges. In a feed-forward network architecture, nodes are organized into an input layer, one or more hidden layers, and an output layer. Each layer is fully connected to the next. Figure~\ref{fig:ann-chromosome} demonstrates the feedfoward ANN structure. Each of the hidden nodes at hidden layer $n$ perform a transcendental function with inputs equal to the outputs of layer $n-1$ each multiplied by the weight associated with the edge through which they are fed from layer $n-1$ to layer $n$. In a radial basis function (RBF) network, the nodes in each hidden layer calculate the distance between their inputs and a centre vector, and pass a weighted distance to a RBF. Nodes in a sigmoid function network perform a similar operation with a sigmoid function in place of the RBF. The output node is typically linear, computing a weighted sum of its inputs. The output of the ANN is a single real value which is interpreted as a classification confidence. ANN computation is dominated by the number of multiplications at the directed edges, proportional to the number of features in the data set and the number of hidden nodes, and by the transcendental functions at the hidden nodes, which are typically exponentiations.  ANN training has been implemented previously on GPU hardware, however these efforts have focused on training through backpropagation \cite{backprop} as opposed to genetic algorithms.

@@

% ############################################################################
\section{Classifier Design} \label{design}
% ############################################################################

% ----------------------------------------------------------------------------
\subsection{ANN Structure Design} \label{ann}
% ----------------------------------------------------------------------------

While our GA-ANN classifier was designed for the 2009 UC San Diego Data Mining Contest data set, it has been designed in order to be capable of adapting to a variety of data sets.  We have developed a flexible ANN topology, allowing for either one or two layers of hidden nodes and an arbitrary number of hidden nodes per layer (within the limits of GPU memory). In addition, each layer of the ANN is capable of performing RBF or sigmoid calculations.

The classifier receives as inputs real-valued features which are standardized prior to processing by the ANN.  The classifier outputs a single real value, in the range of 0 and 1, which represents the confidence with which the pattern is a member of the positive class.  These confidence values can be used for ranking during lift calculations, or for sensitivity or specificity measures with an arbitrary value chosen as a threshold for positive identification of a pattern.  Therefore, the fitness function is not limited to lift at 20\% as presented in this report.

% ----------------------------------------------------------------------------
\subsection{Genetic Algorithm Design} \label{ann}
% ----------------------------------------------------------------------------

For a single-layer ANN, as used in our final design, each candidate ANN is represented by a chromosome which consists of four arrays of real-valued numbers. Edge weights between input nodes and hidden layers, edge weights between hidden and output layers, width values for RBFs at hidden nodes and center values for RBFs at hidden nodes are all encoded, and can take on any real value. Figure~\ref{fig:ann-chromosome} demonstrates the chromosome structure and its relation to ANN structure.  This structure is simple yet powerful; individual edges are 'removed' by the genetic algorithm when their weight is set to zero and nodes are 'removed' by the genetic algorithm when their RBF width value is set to zero.  Thus, while the topology of the ANN is limited to those with $m$ layers and $n$ nodes per layer, the genetic algorithm searches all topologies with $m$ layers and between 0 and $n$ nodes per layer.

@@

% ----------------------------------------------------------------------------
\subsection{Data Preprocessing Steps} \label{preprocessing}
% ----------------------------------------------------------------------------
In order to handle nominal feature values in data sets, we have implemented an algorithm which, given a tab seperated values file, converts nominal feature values in the data set into numeric values.  The algorithm replaces each nominal value in the data set with the logarithm of the probability ratio between the positive and negative classifications observed given the nominal value.  As a result, meaningful numeric values can be extracted from nominal values without domain knowledge relating to the values. This improves on the traditional orthogonal coding method \cite[slide 17]{MLP-NominalCoding} by avoiding allocation of a disproportionate number of input nodes to nominal features.  The logarithm of the ratio of all positive classifications to all negative classifications is attributed to nominal values which do not appear sufficiently in the data set and for values which appear only in blind test data sets; this prevents the classifier from attributing significant values to nominal values for which little or no inference can be done.

@@

When applied to the 2009 UC San Diego Data Mining Contest blind test data set, the resulting classifier produced a lift at 20\% value of 4.031, placing 35th in the world for classification of this data set. Due to the stochastic nature of genetic algorithms and the contest rules which state that contestants may test any number of candidate classifiers, it is likely that repeated entry into this contest would result in a higher ranking. The discrepency between the estimated performance and actual performance on the blind test data set can be attributed to the following:
\begin{itemize}
	\item Discrepency in test set size: the estimated performance measurements were done on sets of approximately 14000 patterns while the blind test data set has approximately 36000 patterns.
	\item Discrepency in class balance: when splitting the data set into training and hold-out data sets, class balance was not enforced for either of the sets. The blind test data set was known to have the same class balance as the given data set. Therefore, it is possible that the hold-out data set used to collect experimental results had a class balance which did not reflect the blind test data set.
	\item Nominal values unique to the blind test data set: It is possible that patterns with nominal values that do not appear in the training data set could cause misclassification, however, the pre-processing step which converts nominal values into numeric values accounts for values which have not been seen previously.
\end{itemize}
